Comments on \(Q\)-Bosons, deformed fermions and their relations with integrable hierarchies

### 2024-04-11 draft.tex
work on the phase model

\address{\noindent 
Instituto de Fíısica Teórica, UNESP-Universidade Estadual Paulista
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
and
Instituto de Física\\ Universidade de S\~ao Paulo\\ 
Rua do Matão Travessa 1371, 05508-090\\ São Paulo, SP. Brazil
}

\email{\texttt{\href{tharaujo@if.usp.br}{tharaujo@if.usp.br}}}

@@

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*
\section{q-Bosons and the Phase Model}
%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*

This section intruduces the \(q\)-boson and phase
models~\cite{Bogoliubov:1992, Bogoliubov:1997soj, Bogoliubov2005,
  Bogoliubov:1997soj, Tsilevich:2006}. We mainly follow the notation
of~\cite{Wheeler:2010vmq}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The Hamiltonian is given by
\begin{equation}
  H = - \frac{1}{2} \sum_{n =0}^M \left(\phi_n^\dagger \phi_{n+1}
  + \phi_n \phi_{n+1}^\dagger \right) + \bar{\mathcal{N}}\; .
\end{equation}
where \(\bar{\mathcal{N}} = \sum_{i=0}^M \mathcal{N}_i\), and we also
impose periodic boundary conditions \(\phi_{M+1} = \phi_0\) and
\(\phi_{M+1}^\dagger = \phi_0^\dagger\).

@@

With these expressions, one can finally build the Bethe states and their dual 
\begin{equation}
  \ket{\Psi(y_1, \dots, y_N)} = \prod_{j=1}^N B(y_j) \ket{\bm{0}}\qquad 
  \bra{\Psi(y_1, \dots, y_N)} = \bra{\bm{0}}\prod_{j=1}^N C(y_j) \; .
\end{equation}
When the set \(\{ y_j \ | \ j =1, \dots , N\}\) satisfies the Bethe equations 
\begin{equation}
  y^{N + M}_i = (-1)^{N-1} \prod_{\substack{j = 1 \\ j \neq i}}^N y_j\; , \qquad i = 1, \dots, N\; , 
\end{equation}
we says that the Bethe states are \emph{on-sheel}. In what follows, we will 
only consider \emph{off-shell} states.

@@

\subsection{q-Bosons}
The Hamiltonian for the q-boson model is
given by
\begin{equation}
  \mathcal{H} = -\frac{1}{2} \sum_{n=0}^M
  \left(b_n^\dagger b_{n+1} + b_n b_{n+1}^\dagger \right) + \bar{\mathcal{N}}\; ,
\end{equation}
where \(\bar{\mathcal{N}} = \sum_{i=0}^M \mathcal{N}_i\), and as in
the case of the phase mode, we impose periodic boundary conditions
\(b_{M+1} = b_0\) and \(b_{M+1}^\dagger = b_0^\dagger\). The set of
operators \(\{b_i, b^\dagger,\mathcal{N}_i\}_{i=0}^M\) form the
\(M+1\) copies of independent q-boson algebras:
\begin{equation}
[\mathcal{N}_i, b_j^\dagger]=\delta_{i,j} b_i^\dagger\; , \quad 
[\mathcal{N}_i, b_j]=-\delta_{i,j}b_i\; , \quad
[B_i, B_j^\dagger]= \delta_{i,j} q^{-2\mathcal{N}_i}  \equiv \delta_{i,j} Q^{\mathcal{N}_i}\; , 
\end{equation}
where we write the deformation parameter as \(Q = q^{-2}\). 
In the limit \(Q\to 1\), the q-bosons are ordinary bosons. 
In the limit \(q\to \infty\), we recover the phase model we discussed above.

\subsubsection{Representation}

\subsubsection{Bethe Ansatz}

@@

\subsection{Bethe Ansatz for the q-boson}

The \(L\)-operator for the q-boson is given by
\begin{equation}
  L_n(\lambda) =
  \begin{pmatrix}
    e^\lambda & \chi B_n^\dagger \\ \chi B_n & e^{-\lambda}
  \end{pmatrix}
\end{equation}
where \(\lambda \in \mathbb{C}\) is the spectral parameter and \(\chi
= \sqrt{1 - q^{-2}}\). From this expression, we have the monodromy matrix
\begin{equation}
  T(\lambda) = \overleftarrow{\prod_{n = 1}^M} L_n(\lambda) =
  \begin{pmatrix}
    A(\lambda) & B(\lambda) \\ C(\lambda) & D(\lambda)
  \end{pmatrix}
\end{equation}
The eigenstates of the Hamiltonian are of the form
\begin{equation}
\ket{ \psi_N(\lambda_1, \dots, \lambda_N)} = \prod_{j=1}^N B(\lambda_j) \ket{0}\; , 
\end{equation}
where the parameters \(\{\lambda_i\}_{i=1}^N\) satisfies the Bethe
Algebraic Equation
\begin{equation}
  e^{2M\lambda_j} = \prod_{k\neq j}^N \frac{f(\lambda_k, \lambda_j)}{f(\lambda_j, \lambda_k)}\quad \textrm{with} \quad
  f(\lambda, \mu) = \frac{\sinh(\lambda-\mu+\gamma)}{ \sinh(\lambda- \mu)}
  \quad j = 1,\dots \ldots , N\; .
\end{equation}
where

\subsection{Bethe Ansatz for the phase model}

Taking the limit \(\gamma\to \infty\) we obtain the ABA for the phase model.
In particular, we find
\begin{equation}
  L_n(\lambda) =
  \begin{pmatrix}
    e^{ip/2} & \phi_n^\dagger \\ \phi_n & e^{-ip/2}
  \end{pmatrix}
\end{equation}
where we write \(\lambda = i p/2\). The Algebraic Bethe Equations are
\begin{equation}
\exp \left(i (M + N) p_j \right) = (-1)^{N-1} \exp \left( i \sum_{k=1}^N p_k \right)\quad j = 1, \dots, N\; ,
\end{equation}
that can be solved explicitly as
\begin{equation}
  \label{}
p_j = \frac{ 2 \pi I_j + \sum_{k=1}^N p_k}{M+N}\; , 
\end{equation}
where \(i_j \in \mathbb{z}\) ( \( + 1/2\)) if \(n\) is odd (even).

In what follows, we consider the transformation \(x^{-1} = e^{ip}\), see~\cite{Wheeler:2010vmq},
in such a way that the \(L\)-operator becomes
\begin{equation}
  L_{an}(\lambda) =
  \begin{pmatrix}
    x^{-1/2} & \phi_n^\dagger \\ \phi_n & x^{1/2} 
  \end{pmatrix}_a
\end{equation}

\subsubsection{Bethe states}

Now we also consider the Bethe states (and their dual) 
\begin{equation}
\ket{\Psi(\bm{x})} = \prod_{j=1}^N B(x_j) \ket{\bm{0}}\quad \; ,
\bra{\Psi(\bm{x})} = \bra{\bm{0}}\prod_{j=1}^N C(x_j) \; ,
\end{equation}
where we have written \(\bm{x} = (x_1, \dots, x_N)\), and the monodromy matrix
is given by
\begin{equation}
  T_a(x) = \prod_{j=1}^M L_{a j}(x) = 
  \begin{pmatrix}
    A(x) & B(x) \\
    C(x) & D(x)
  \end{pmatrix}_a \; .
\end{equation}
Folowing~\cite{Wheeler:2010vmq}, we define \(\mathbb{B} = x^{M/2} B(x)\) and 
\(\mathbb{C} = x^{M/2} C(x^{-1})\).

\subsection{Scalar product}

>>>>

\address{\noindent 
Instituto de Física Teórica, UNESP-Universidade Estadual Paulista
R. Dr. Bento T. Ferraz 271, Bl. II, Sao Paulo 01140-070, SP, Brazil\\
and
Instituto de Física\\ Universidade de S\~ao Paulo\\ 
Rua do Matão Travessa 1371, 05508-090\\ São Paulo, SP. Brazil
}

\email{\texttt{\href{tr.araujo@unesp.br}{tr.araujo@unesp.br}}}

@@

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*
\section{Phase Model and q-Bosons}
%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*

This section introduces the \(q\)-boson and phase
models~\cite{Bogoliubov:1992, Bogoliubov:1997soj, Bogoliubov2005,
  Bogoliubov:1997soj, Tsilevich:2006}, and consists of a general
review of the literature. Consequently, there is nothing particularly
original, except its presentation.  In spite of that, I believe that
it has some pedagoginal value, and a good collection of references.
Here we follow some conventions of~\cite{Wheeler:2010vmq}, but with a
presentation that is closer to~\cite{Tsilevich:2006}.

@@

The Hamiltonian is given by
\begin{equation}
  H = - \frac{1}{2} \sum_{n =0}^M \left(\phi_n^\dagger \phi_{n+1}
  + \phi_n \phi_{n+1}^\dagger \right) + \bar{\mathcal{N}}\; .
\end{equation}
where \(\bar{\mathcal{N}} = \sum_{i=0}^M \mathcal{N}_i\), and
periodic boundary conditions \(\phi_{M+1} = \phi_0\) and
\(\phi_{M+1}^\dagger = \phi_0^\dagger\).

@@

With these expressions, one can finally build the Bethe states and their dual 
\begin{equation}
  \ket{\widehat{\Psi}(y_1, \dots, y_N)} = \prod_{j=1}^N B(y_j) \ket{\bm{0}}\qquad 
  \bra{\widehat{\Psi}(y_1, \dots, y_N)} = \bra{\bm{0}}\prod_{j=1}^N C(y_j) \; .
\end{equation}
When the set \(\{ y_j \ | \ j =1, \dots , N\}\) satisfies the Bethe equations 
\begin{equation}
  y^{N + M}_i = (-1)^{N-1} \prod_{\substack{j = 1 \\ j \neq i}}^N y_j\; , \qquad i = 1, \dots, N\; , 
\end{equation}
we says that the Bethe states are \emph{on-sheel}. In what follows, we will 
only consider \emph{off-shell} states.

\subsubsection{Coordinate expansion}
From the map defined in~\cite{Wheeler:2010vmq}, we can associate to
each state \(\ket{\lambda}\), a Schur function \(s_\lambda(\vec{x})\),
where the dummy coordinates are \(\vec{x} = (x_1, \dots, x_N)\), with
\(N = \sum_{j=0}^M n_j\). Therefore, we can work in the ring of
symmetric functions \(\Lambda_M\), with basis given by Schur functions
\(s_\lambda\) and Young diagrams such that \(\max(\lambda_1) = M\),
and \(\max(\ell(\lambda)) = N\).

That also means that the operator \(B\) should also be realized in the
space \(\Lambda_M\), see~\cite{Tsilevich:2006}. The operators
\(\phi^\dagger_j\) and \(\phi_j\) are, respectively, \(1\)-particle
creation and annihilation at the site \(j\). In terms of partitions,
we can think of them as insertion and removal of a row of length
\(j\). That is, given the state \(\ket{\lambda} =\ket{1^{n_1} 2^{n_2}\cdots}\), 
we have
\begin{equation}
\begin{split}
  \phi^\dagger_j \ket{\lambda} & = \ket{1^{n_1} 2^{n_2}\cdots j^{j+1}\cdots} \equiv \ket{\lambda \oplus (j)}\\
  \phi_j \ket{\lambda} & = \ket{1^{n_1} 2^{n_2}\cdots j^{j-1}\cdots}
  \equiv \ket{\lambda \ominus (j)}\; , 
\end{split}
\end{equation}
Once again, we see that the case \(j=0\) is meaningless in the
partition construction. Given two partitions \(\lambda\) and \(\mu\),
we say that \(\lambda\) interlaces \(\mu\) if \(\lambda_i \geq \mu_i \geq \lambda_{i+1}\), 
and we write \(\lambda\succ \mu\). Therefore, it is easy to see that the partition 
\(\lambda\oplus (j)\) interlaces \(\lambda\), and \(\lambda\)
interlaces \(\lambda \ominus (j)\). We write this condition as 
\(\lambda \oplus (j) \succ \lambda \succ \lambda \ominus (j)\).

Now, we can use the explicit expression for the \(L\)-matrix to build
the monodromy matrix, and extract the operator \(B(x)\) as
\begin{subequations}
\begin{equation}
  B(x)_a = T_a(x)_{12} = \sum_{j_1, \dots, j_M = 1,2} L_{a j_M}(x) L_{j_M j_{M-1}}(x)\cdots  L_{j_1 2}(x)\; ,
\end{equation}
and it is easy to see that it has the form  
\begin{equation}
  B(x) \equiv x^{-M/2} \mathbb{B}(x) = x^{-M/2} \sum_{k=0}^M x^k \mathcal{B}_k \; .
\end{equation}
The problem now is to find the action of the operators \(\mathcal{B}_k\) on
states \(\ket{\lambda}\).
\end{subequations}
Let us consider some simple cases.

\textbullet \ M=0: We have just one site in the  
chain, and all particles site at the site \(0\), as we have discussed many times, 
does not contribute to the partition representation.

\textbullet \ M=1: Now we have 1 row Young diagrams \(\lambda = (1^{n_1})\), where each row
corresponds to a boson. It is easy to show that 
\begin{equation}
 \mathbb{B}_{M=1}(x) =  x \phi_1^\dagger + \phi_0^\dagger \; ,
\end{equation}
therefore, \(\mathcal{B}_0 = \phi_0^\dagger\) and  \(\mathcal{B}_1 = \phi_1^\dagger\). 
From this expression, we have 
\begin{equation}
  \mathcal{B}_0\ket{(1^{n_1})} = \ket{(1^{n_1})} \qquad  
  \mathcal{B}_1\ket{(1^{n_1})} = \ket{(1^{n_1+1})} \; .
\end{equation}
All in all, we conclude that 
\begin{equation}
 \mathbb{B}_{M=1}(x)\ket{\lambda}  =  \ket{\lambda} + x\ket{\lambda \oplus (1)} \;.
\end{equation}

\textbullet \ M=2: Now we have 2 row Young diagrams \(\lambda = (1^{n_1}2^{n_2})\).
Now, one can show that
\begin{equation}
  \mathbb{B}_{M=2}(x) = x^2 \phi_2^\dagger + x \left( \phi_1^\dagger
  + \phi_2^\dagger \phi_1 \phi_0^\dagger \right) + \phi_0^\dagger \; ,
\end{equation}
therefore,
\begin{equation}
\mathcal{B}_0 = \phi_0^\dagger \qquad 
\mathcal{B}_1 = \phi_1^\dagger + \phi_2^\dagger \phi_1 \phi_0^\dagger \qquad 
\mathcal{B}_2 = \phi_2^\dagger \; .
\end{equation}

Consequently. we find
\begin{subequations}
\begin{equation}
\begin{split}
  \mathcal{B}_0\ket{(1^{n_1}2^{n_2})} & = \ket{(1^{n_1} 2^{n_2})} \qquad  
  \mathcal{B}_1\ket{(1^{n_1}2^{n_2})} = \ket{(1^{n_1+1} 2^{n_2})} + \ket{(1^{n_1-1} 2^{n_2 + 1})} \\
  \mathcal{B}_2\ket{(1^{n_1}2^{n_2})} & = \ket{(1^{n_1}2^{n_2+1})}\; .
\end{split}
\end{equation}
\end{subequations}
Observe that the action of the operator \(\mathbb{B}_k(x)\) on \(\ket{\lambda}\)
generates diagrams \(\ket{\mu}\) such that \(k = |\mu| - |\lambda|\). 
\begin{equation}
  \mathbb{B}_{M=2}(x) \ket{\lambda}  =
  \sum_{\lambda \prec \mu \subseteq [2,\ell(\lambda) + 1]} x^{|\mu|-|\lambda|} \ket{\mu}\; .
\end{equation}

One can keep doing this game, and we reach to the general conclusion that 
\begin{equation}
  \mathbb{B}(x) \ket{\lambda}  =
  \sum_{\lambda \prec \mu \subseteq [M,\ell(\lambda) + 1]} x^{|\mu|-|\lambda|} \ket{\mu}\; .
\end{equation}

Therefore, we conclude that the action of the operators
\(\mathcal{B}_k\) on \(\ket{\lambda}\) generates diagrams
\(\ket{\mu}\) that are obtained from \(\lambda\) by adding \(k\) boxes,
not in different columns. From the viewpoint of symmetric functions, this rule
is exactly the Pieri's rule~\cite{Macdonald:1998}, that states that
we have that
\begin{equation}
  s_{(k)} s_\lambda = \sum_{\mu \succ \lambda} s_\mu\;.
\end{equation}

Therefore, we have the map from the spin chain Hilbert space to the
ring of symmetric functions given by \(\imath : \mathcal{B}_k \mapsto
s_{(k)}\) where \(s_{(k)}\) is the Schur function with 1 row of length
\(k\).

@@

\subsection{q-Bosons}
The Hamiltonian for the q-boson model is
given by
\begin{equation}
  \mathcal{H} = -\frac{1}{2} \sum_{n=0}^M
  \left(b_n^\dagger b_{n+1} + b_n b_{n+1}^\dagger \right) + \bar{\mathcal{N}}\; ,
\end{equation}
where \(\bar{\mathcal{N}} = \sum_{i=0}^M \mathcal{N}_i\), and as in
the case of the phase mode, we impose periodic boundary conditions
\(b_{M+1} = b_0\) and \(b_{M+1}^\dagger = b_0^\dagger\). The set of
operators \(\{b_i, b^\dagger,\mathcal{N}_i\}_{i=0}^M\) form the
\(M+1\) copies of independent q-boson algebras:
\begin{equation}
[\mathcal{N}_i, b_j^\dagger]=\delta_{i,j} b_i^\dagger\; , \quad 
[\mathcal{N}_i, b_j]=-\delta_{i,j}b_i\; , \quad
[B_i, B_j^\dagger]= \delta_{i,j} q^{-2\mathcal{N}_i}  \equiv \delta_{i,j} Q^{\mathcal{N}_i}\; , 
\end{equation}
where we write the deformation parameter as \(Q = q^{-2}\). 
In the limit \(Q\to 1\), the q-bosons are ordinary bosons, and 
the limit \(Q\to 0\) (\(q\to \infty\)) gives the phase model we have discussed above.

\subsubsection{Representation}
We build the representation of the q-boson algebra with some modifications of the phase 
model we have just discussed. We suppose that the action of the operators \(\{b_i, b_i^\dagger\}\) 
satisfies 
\begin{subequations}
\begin{alignat}{3}
    & b_0\ket{n_0} =(1 - \delta_{0, n_0}) \frac{1 - Q^{n_0}}{(1 - Q)^{1/2}}\ket{n_0 - 1}
    &\hspace{0.5cm}& b_0^\dagger \ket{n_0} =  \frac{1}{(1 - Q)^{1/2}}\ket{n_0 + 1}  \\
    & b_i\ket{n_i} = \frac{(1 - \delta_{0, n_i})}{(1 - Q)^{1/2}}\ket{n_i - 1}
    && b_i^\dagger \ket{n_0} =  \frac{(1 - Q^{n_i+1})}{(1 - Q)^{1/2}}\ket{n_0 + 1}\quad i\neq 0
\end{alignat}
\end{subequations}
where the modes \(\{b_0, b_0^\dagger \}\) satisfies a slightly different relation.

\subsubsection{Bethe Ansatz}
The \(L\)-operator for the q-boson is given by
\begin{equation}
  L_n(x, Q) =
  \begin{pmatrix}
    x^{-1/2} & (1 - Q)^{1/2} b_n^\dagger \\ (1 - Q)^{1/2} b_n & x^{1/2}
  \end{pmatrix}\; .
\end{equation}
The monodromy matrix is 
\begin{equation}
  T_a(x,Q) = L_{am}(x, q)  \dots  L_{a0}(x, q) = 
  \begin{pmatrix}
    A(x) & B(x) \\ C(x) & D(x)
  \end{pmatrix}\; .
\end{equation}

The eigenstates of the Hamiltonian are of the form
\begin{equation}
  \ket{\Psi(y_1, \dots, y_N; Q)} = \prod_{j=1}^N B(y_j, Q) \ket{\bm{0}}\qquad 
  \bra{\Psi(y_1, \dots, y_N; Q)} = \bra{\bm{0}}\prod_{j=1}^N C(y_j, Q) \; .
\end{equation}
When the set \(\{ y_j \ | \ j =1, \dots , N\}\) satisfies the Bethe equations 
\begin{equation}
  y^{N + M}_i =\prod_{\substack{j = 1 \\ j \neq i}}^N\frac{Q y_i - y_j}{y_i - Q y_j}\; , \qquad i = 1, \dots, N\; , 
\end{equation}
Once again, we will consider \emph{off-shell} Bethe states.

@@

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*
\section{Toda tau functions and the phase model}
%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*
